Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2015-04-28 doc/ApTranscriptome_manuscript.Rmd
Edits in response to reviewer comments.

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: molecol.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated a *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation.

@@

We collected a single colony of each of the ant species *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and *Aphaenogaster carolinensis*, part of the *A. rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012. Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University). Colonies of these species were maintained in common conditions at 25°C for 6 months prior to experimentation. We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W). Following the protocol of Warren and Chick [-@warren2013], we tested the upper and lower critical thermal limits for 5 ants from each of these colonies.

@@

To characterize the thermal reactionome, we haphazardly collected 10 ants from each colony at the same time on 12 days. Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013]. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$

to each transcript. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005]. This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.

@@

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Fig. S1) indicating that these differences are not simply due to neutral processes. The *enhanced defense* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species. The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species. For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A. carolinensis* for low temperatures, *A. picea* for high temperatures) and testing whether those genes displayed the same response profile in the better-adapted species. As with the overall results, the patterns of association differed for the two temperature extremes. Genes upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from all other expression categories in *A. picea*, suggesting shared response pathways (Fig. 1A). In contrast, genes upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) were only weakly associated with this same category in *A. carolinensis* (Fig. 1B). Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig. 1B).

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A. carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2A). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2A).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig. 2). There was greater inertia of gene expression in response to high and low temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures **Intermediate** transcripts from *A. picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A. carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 2B).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the midpoint temperature (19.5°C). If stress-response genes have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A. picea* (Fig. 3A) or transcripts upregulated at low temperatures in *A. carolinensis* (Fig. 3B). In fact, for both comparisons we found a weakly positive relationship.

In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in far-right column of Fig. 1A). Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A. carolinensis*.

@@

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 2A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

In contrast to cold tolerance, the enhanced upper thermal limit in *A. picea* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. Similarly, significantly higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 2B-D). This difference between species is also evident among candidate genes known to be involved in stress response. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress. These differences are in line with expectations that *A. carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions. In fact, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

@@

Given the differences in the patterns of thermal responsiveness between species (Fig. 1, 2), it is worth noting a number of similarities. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012]. The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role. Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 2). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

@@

Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in , expression type categories or critical temperature thresholds as we do here (Fig. 1, Fig. S1). Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression, as opposed to identifying lists of differentially-expressed genes, can provide meaningful insight on the process of evolution.

As with all *de novo* transcriptomes, our work has some caveats. First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness.

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 8% of their transcriptomes are thermally responsive. For both species, more transcripts responded to low than high temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have more limited genetic accomodation to high than low temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626. The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive). The reproducible and version-controlled script for data analysis is available from GitHub (https://github.com/johnstantongeddes/ApTranscriptome).

@@

**Figure 1**. Deviations from expected numbers of transcripts in matched observations of transcript expression type. The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (blue = excess, red = deficit). The dashed boxes indicate the matched (e.g. **High** - **High**). (A) **Low** and **Bimodal** expression types for *A. carolinensis* showing bias away from all other categories in *A. picea*. (B) **High** and **Bimodal** expression types for *A. picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**). The complete set of matched observations is shown in Fig. S1.

**Figure 2**. Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*.

>>>>

---
title: Thermal reactionomes reveal divergent responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: molecol.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) and characterized the reaction norm for each gene using a custom analysis. We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adaptation.

@@

We collected a single colony of *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and a single colony of *Aphaenogaster carolinensis*, part of the *A. rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012. These locations are central to each species distributions. Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University). Colonies of these species were maintained in common conditions at 25°C for 6 months prior to experimentation. We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W). We tested the upper and lower critical thermal limits for 5 ants from each of these colonies using a ramp of 1° C per minute, starting at 30° C, and recorded the temperature at which the ants were no longer able to right themselves [@warren2013].

@@

To characterize the thermal reactionome, we haphazardly collected three ants from each 2012 colony at the same time on 12 days. Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temperature) + \beta_3(temperature^2) + \beta_4(species * temperature) + \beta_5(species * temperature^2) + \epsilon $$

with to each transcript. Temperature and species were both fixed effects, with a quadratic term included for temperature. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005]. This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.

@@

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Fig. S1) indicating that these differences are not simply due to neutral processes. The *enhanced defense* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species. The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species. For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A. carolinensis* for low temperatures, *A. picea* for high temperatures) and testing whether those genes displayed the same response profile in the better-adapted species. As with the overall results, the patterns of association differed for the two temperature extremes. Genes upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from all other expression categories in *A. picea*, suggesting shared response pathways (Fig. 2A). In contrast, genes upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) were only weakly associated with this same category in *A. carolinensis* (Fig. 2B). Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig. 2B).

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A. carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 3A). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 3A).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig. 3). There was greater inertia of gene expression in response to high and low temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures **Intermediate** transcripts from *A. picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A. carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 3B).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the midpoint temperature (19.5°C). If stress-response genes have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A. picea* (Fig. 4A) or transcripts upregulated at low temperatures in *A. carolinensis* (Fig. 4B). In fact, for both comparisons we found a weakly positive relationship.

In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (far-right column of Fig. 2B). Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A. carolinensis*.

@@

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 3A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

In contrast to cold tolerance, the enhanced upper thermal limit in *A. picea* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. Similarly, significantly higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 3B-D). This difference between species is also evident among candidate genes known to be involved in stress response. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress. These differences are in line with expectations that *A. carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions. In fact, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

@@

Given the differences in the patterns of thermal responsiveness between species (Fig. 2, 2), it is worth noting a number of similarities. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role. Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 3). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

@@

Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in , expression type categories or critical temperature thresholds as we do here (Fig. 2, Fig. S1). Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression, as opposed to identifying lists of differentially-expressed genes, can provide meaningful insight on the process of evolution.

As with all *de novo* transcriptomes, our work has some caveats. First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness. In particular, functional protein levels cannot be expected to be directly linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability (...citations…).

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 8% of their transcriptomes are thermally responsive. For both species, more transcripts responded to low than high temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have more limited genetic accomodation to high than low temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing expression of transcripts related to metabolism, heat shock proteins and other protective molecules, whereas *A. carolinensis* decreases expression of transcripts related to metabolism and likely relies on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626. The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-39, and hf113-40. The reproducible and version-controlled script for data analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).

@@

**Figure 1**. Cartoon figure showing the predicted results for the deviations from the expected number of transcripts in each expression type. The color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. The (A) *enhanced defense* hypothesis assumes that temperature adapation uses the same mechanisms, and thus that there is a strong overlap of response (e.g. excess in shared response types between species). In contrast, the (B) *tolerance* hypothesis predicts that responsive transcripts in one species will become non-responsive or shift expression type in the other species (e.g. excess in non-shared response). Expression types are defined in the text and Table 1. We show only the results for *A. carolinensis* in response to low temperatures as the pattern would be similar *A. picea* in response to high temperatures.
